Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-10-07 MainText.Rmd
edits

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) 
<!--***DAN: Shyamolina, I do not know what the parenthesis refers to -->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annal survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction. 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

@@

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail dependence in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite dependence in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail dependence
with positive Kendall's correlation ($\tau = 0.6$). 
Similarly, we generated extreme left- (or right-) tail dependent matrix M 
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula.

We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we feed pairwise entries from each row of matrix $M$ as the vital rates 
(fecundity $f_a$ and survival rates $s_e$) 
of the stochastic projrction matrix in model (\ref{eq_age_str}).

@@

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
dependence in left and right-tail.

<!--say how you set ext. threshold = Pe/10-->
We started with $n$ patches of initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

The models stated in Table \ref{tab_model_summary} had one common 
feature: they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level, say, $r_c$. Models showed 
overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$ when $r$ was beyond $r_c$. 
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond Hopf-bifurcation point, model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.

>>>>

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) 
<!--***DAN: Shyamolina, I do not know what the parenthesis refers to -->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annual survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction. 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

@@

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail dependence in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite dependence in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail dependence
with positive Kendall's correlation ($\tau = 0.6$).

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we feed pairwise entries from each row of matrix $M$ as the vital rates 
(fecundity $f_a$ and survival rates $s_e$) 
of the stochastic projrction matrix in model (\ref{eq_age_str}).

@@

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
dependence in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

The models stated in Table \ref{tab_model_summary} had one common 
feature: they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level (say, $r_c$) and overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$ when $r$ was beyond $r_c$. 
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond Hopf-bifurcation point, model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.
